### 2024-04-25 parts/correction.tex
knowledges

\AP The key lemma of this section is \cref{lem:correct-to-n-rat}, which is
proved by induction on the \kl{maximal monomials} of a polynomial $P$. The core
combinatorial argument allowing to perform the induction is
\cref{derivation-stabilises-correct:lem}, which is proved by a straightforward
binomial expansion. Using this lemma, the induction is performed as follows:
after decomposing a polynomial $P$ into a sum of \kl{maximal monomials} $P_1$
and the rest $P_2$, notice that for all $K \in \Nat$, $\translate{K}(P) = P_1 +
\Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
is a \kl{$\Nat$-rational polynomial}, furthermore, we can use the induction
hypothesis on $\Diff{K}{P_1} + \translate{k}(P_2)$ since it has lower degree
monomials and is in $\CorrectPoly$, finally we can also apply the induction
hypothesis to $\restr{P}{\nu}$ where $\nu$ is a partial function from $\vec{X}$
to $\set{0, \dots, K}$ fixing at least one variable. It is then just a matter
of combining these functions to obtain a \kl{$\Nat$-polyregular function} that
\kl{represents} $P$. Notice that in \cref{lem:correct-to-n-rat}, polynomials
can be represented using \kl{star-free $\Nat$-polyregular functions}, which is
a stronger statement than the one in \cref{karh:thm}, but is in line with the
idea that polynomials are not periodic functions (for instance, in
\cref{zsf-nsf:conjecture}).

>>>>

\AP The key lemma of this section is \cref{lem:correct-to-n-rat}, which is
proved by induction on the \kl{maximal monomials} of a polynomial $P$. The core
combinatorial argument allowing to perform the induction is
\cref{derivation-stabilises-correct:lem}, which is proved by a straightforward
binomial expansion. Using this lemma, the induction is performed as follows:
after decomposing a polynomial $P$ into a sum of \kl{maximal monomials} $P_1$
and the rest $P_2$, notice that for all $K \in \Nat$, $\translate{K}(P) = P_1 +
\Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
is a \kl{$\Nat$-rational polynomial}. Furthermore, we can use the induction
hypothesis on $\Diff{K}{P_1} + \translate{k}(P_2)$ since it has lower degree
monomials and is in $\CorrectPoly$, finally we can also apply the induction
hypothesis to $\restr{P}{\nu}$ where $\nu$ is a partial function from $\vec{X}$
to $\set{0, \dots, K}$ fixing at least one variable. It is then just a matter
of combining these functions to obtain a \kl{$\Nat$-polyregular function} that
\kl{represents} $P$. Notice that in \cref{lem:correct-to-n-rat}, polynomials
can be represented using \kl{star-free $\Nat$-polyregular functions}, which is
a stronger statement than the one in \cref{karh:thm}, but is in line with the
idea that polynomials are not periodic functions (this will be made precise in
\cref{zsf-nsf:conjecture}).

### 2024-04-25 parts/intro.tex
decidability of decision
knowledges

\AP In order to approach the decidability of decision problems on
\kl{polyregular functions}, restricting the output alphabet to a single letter
has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
single-letter output alphabets is that the particular order in which letters
are produced by the machine is irrelevant. This allows to obtain results that
would otherwise fail because of this ordering, such as the so-called
\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
output} in the case of \kl{polyregular functions} can often be lifted to
outputs in the free commutative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
really about enforcing commutativity on the output. However, even with the
restriction to single letter output alphabets, the decidability of
\kl{star-free polyregular functions} among \kl{polyregular functions} remains
an open problem \cite[Conjecture 7.61]{DOUE23}.

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
authors introduced the class of \kl{$\Rel$-polyregular functions}
($\intro*\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
negative outputs. This remains a subclass of \kl{rational series}, more
precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
this class of functions the aperiodic variant $\intro*\ZSF$ has been shown to
be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
we only refer to the one applying to rational series of polynomial growths
introduced in \cite{CDTL23}.

>>>>

\AP In order to approach the decision problems on
\kl{polyregular functions}, restricting the output alphabet to a single letter
has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
single-letter output alphabets is that the particular order in which letters
are produced by the machine is irrelevant. This allows to obtain results that
would otherwise fail because of this ordering, such as the so-called
\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
output} in the case of \kl{polyregular functions} can often be lifted to
outputs in the free commutative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
really about enforcing commutativity on the output. However, even with the
restriction to single letter output alphabets, the decidability of
\kl{star-free polyregular functions} among \kl{polyregular functions} remains
an open problem \cite[Conjecture 7.61]{DOUE23}.

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
authors introduced the class of \kl{$\Rel$-polyregular functions}
($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
negative outputs. This remains a subclass of \kl{rational series}, more
precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
this class of functions the aperiodic variant $\ZSF$ has been shown to
be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
we only refer to the one applying to rational series of polynomial growths
introduced in \cite{CDTL23}.

### 2024-04-25 parts/prelim.tex
knowledges

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.

>>>>

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes of
\intro{$\Rel $-polyregular functions}
$\intro*\ZPoly[k]$,
\intro{$\Nat $-polyregular functions}
$\intro*\NPoly[k]$,
\intro{star-free $\Rel $-polyregular functions}
$\intro*\ZSF[k]$, and
\intro{star-free $\Nat $-polyregular functions}
$\intro*\NSF[k]$, 
as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.
